---
title: "Classification Methods Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification Methods Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ada.classification)
```

#### k-means classification and the kmeansPlus() function

##### </br > K-means is a classification procedure designed to separate data into a specified number of clusters or groups such that differences between members of each group are minimized. In this context, a number of centers are randomly chosen (or specifically assigned, depending on what is passed to the 'centers' argument of the function) and clusters are formed by minimizing the sum of squares between data points and their nearest center. This process is iterated until the best fitting clusters have been assigned. Looking at the within-cluster sum of squares as a proportion of the the total sum of squares for all data points can provide a measure of how distinctive the resultant clusters are. The kmeansPlus() function seeks to aid users in implementing the k-means procedure and interpretting its results.


##### </br > Below we explore k-means clustering using the kmeansPlus() function on the 'diabetes' dataset. when 'plot' is set to FALSE, the output is identical to what one would obtain by running stats::kmeans(). The only major difference is the 'exclude' argument, which allows the user to name variables in the dataset to be excluded from the k-means procedure. In this case we set k = 2 and exclude the 'Outcome' variable. This produces two clusters, which account for all variables except the one indicating whether or not the patient was diagnosed with diabetes.

```{r}

set.seed(123)

kmeansPlus(diabetes, 2, exclude = "Outcome", plot = FALSE)

```

##### We can see from the value of 55.7% for between_SS / total_SS (sum of squares within groups / sum of squares of data points as a whole) that the clusters produced when k is set to 2 are not very distinct. That is, we were not able to minimize the extent of difference betweeen members of the same group as much as we would like. Below we run the function again, but this time leaving 'plot' set to the default value of TRUE. We now must specify an x and y variable to be plotted.

```{r}

kmeansPlus(diabetes, 2, exclude = "Outcome", x = "Age", y = "BMI")

```


##### The output in the console remains the same, but there is now a plot as well. The plot allows us to vizualize the distinctiveness of the clusters assigned by our implementation of k-means, but only for the two variables provided as inputs to the 'x' and 'y' arguments. As we can see, the clusters assigned do not seem very meaningful for the intersection of Age and BMI. Below we try out a few more combinations of variables, this time setting 'full.output' to FALSE (since we are only interested in the plots, and have no need to see the same kmeans output each time we create one).

```{r}

p1 <- kmeansPlus(diabetes, 2, exclude = "Outcome", full.output = FALSE, x = "Glucose", y = "BMI")
p2 <- kmeansPlus(diabetes, 2, exclude = "Outcome", full.output = FALSE, x = "BloodPressure", y = "Age")
p3 <- kmeansPlus(diabetes, 2, exclude = "Outcome", full.output = FALSE, x = "Insulin", y = "SkinThickness")
p4 <- kmeansPlus(diabetes, 2, exclude = "Outcome", full.output = FALSE, x = "Age", y = "Insulin")

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)

```

##### Based on this preliminary exploration of different combinations of variables, Inslulin is the only variable for which the clusters produced by our k-means procedure are distinctive. This is a meaningful finding given the nature of the dataset, although it is possible that this is simply a querk caused by the fact that the 'Insulin' variable seems to contain a large number of zero values. Regardless, this has demonstrated the use of simple scatterplots to aid in the interpretation of k-means outputs. We can conclude that 2 may not be a very effective k value, based on both the middling sum of squares ratio of 55.7% and on the plots, which only reflect distinctive groups in cases where the 'Insulin' variable is included on one of the axes.


##### </br > We can now set the 'k' argument to 3 instead of 2 to see how this changes the results. We will find a much better sum of squares ratio of 74.9%, indicating that this may be a more effective value for k. The plots, however, will continue to reveal that 'Insulin' is the variable responsible for most of the distinctive clustering. This would agree with what is shown in the full k-means output, where the cluster means of the two (or three) clusters vary more dramatically for insulin than they do for the other variables.

```{r}

kmeansPlus(diabetes, 3, exclude = "Outcome", x = "BMI", y = "BloodPressure")

```




#### </br > Finally, we see what happens when both 'full.output' and 'plot' are set to FALSE. The output is a tibble that is simply the original input dataset with a new column added, indicating the cluster to which each case was assigned. This is just a simple way of taking the cluster vector from the output of stats::kmeans() and adding it to the original dataset as a new column. This is mainly for when the user wants to perform subsequent operations using the groupings produced by k-means. This would allow one to, for example, observe whether there is meaningful correspondence between the clusters assigned when k=2 and the values in the 'Outcome' column, which also separate the cases into one of two classes (those who were diagnosed with diabetes and those who were not). Given that Insulin is the most influential variable in the classifications produced by k-means when k=2, it would be interesting to explore this question further (and the tibble produced below would be a good place to start!).

```{r}
# Viewing the original dataset. We exclude the first two columns just so that we can see the last few columns when it prints.
diabetes[, 3:ncol(diabetes)]


# Running a simple implementation of kmeansPlus and viewing the result
result <- kmeansPlus(diabetes, 2, exclude = "Outcome", full.output = FALSE, plot = FALSE)
result[, 3:ncol(result)]
```









